---
title: "Stats / Fact-checks for 'why US needs a $17 minimum wage'"
format: html
theme: cosmo
date: "`r Sys.Date()`"
date-format: long
execute:
  echo: false
  output: false
---

\captionsetup[table]{labelformat=empty}
\vspace{-3em}
```{r setup}
library(tidyverse)
library(targets)
library(bea.R)

microdata_results <- tar_read(prep_acs_data)
```

```{r mw_sim results}
cpi_deflator <- 305.535 / 344.789

results <- microdata_results %>% 
  mutate(
    weight = perwt6,
    d_affected = direct6 == 1,
    i_affected = indirect6 == 1,
    affected = d_affected == 1 | i_affected == 1,
    ann_wage_change = d_annual_inc6
  ) %>% 
  # getting a weird haven/vctrs error here
  haven::zap_labels() %>% 
  mutate(
    poverty = povstat == 0,
    black = racec == 2,
    hispanic = racec == 3,
    black_hispanic = black == 1 | hispanic == 1,
    black_hispanic_female = (black == 1 | hispanic == 1) & female == 1,
    prime_age = age >= 25 & age <= 54,
    teen = age <= 19,
    full_time = hourc == 2,
    some_college = edc >= 3,
    parent = childc <= 2
  )

stats_by_group <- function(group) {
  counts_shares <- results %>% 
    group_by(group = .data[[group]]) %>% 
    summarize(
      count_affected = sum(affected * weight),
      share_affected = weighted.mean(affected, w = weight)
    ) %>% 
    mutate(group_share_affected = count_affected / sum(count_affected)) %>% 
    filter(group == 1) %>% 
    mutate(group := {{group}})
  
  wage_increases <- results %>% 
    filter(.data[[group]] == 1 & affected == 1) %>% 
    summarize(
      wage_change_avg_ann = weighted.mean(ann_wage_change, w = weight) * cpi_deflator,
      wage_change_total_ann = sum(ann_wage_change * weight) * cpi_deflator
    ) %>% 
    mutate(group := {{group}})
  
  counts_shares %>% 
    full_join(wage_increases, by = "group")
}

groups <- c(
  "all",
  "poverty",
  "black",
  "black_hispanic",
  "black_hispanic_female",
  "hispanic",
  "prime_age",
  "teen",
  "full_time",
  "some_college",
  "parent",
  "female"
)
group_stats <- map(groups, stats_by_group) %>% 
  list_rbind() %>% 
  mutate(across(
    matches("share"), 
    ~ scales::label_percent(accuracy = 1)(.x)
  )) %>%
  mutate(wage_change_total_ann = wage_change_total_ann / 10^9) %>% 
  mutate(wage_change_avg_ann = round(wage_change_avg_ann / 100) * 100) %>% 
  mutate(across(
    matches("^wage_change"), 
    ~ scales::label_dollar(accuracy = 1)(.x)
  )) %>% 
  mutate(count_affected = scales::label_comma(accuracy = 1)(
    count_affected / 10^6
  ))


extract_stat <- function(group, stat) {
  group_stats %>% 
    filter(.data[["group"]] == {{group}}) %>% 
    pull(.data[[stat]])
}
```

```{r productivity}
bea_key <- Sys.getenv("BEA_API_KEY")

bea_grab <- function(tablename) {
  bea_specs <- list(
    'UserID' = bea_key,
    'Method' = 'GetData',
    'datasetname' = 'NIPA',
    'TableName' = tablename,
    'Frequency' = 'A',
    'Year' = 'X'
  )
  beaGet(bea_specs, asWide = FALSE) %>% 
    as_tibble()
}

bea_ndp <- bea_grab("T10705") %>% 
  filter(LineNumber == 14) %>% 
  transmute(year = as.numeric(TimePeriod), ndp = DataValue)

bls_hours <- read_csv("inputs_raw/total-economy-hours-employment.csv") %>% 
  filter(
    Component == "Total U.S. economy",
    Basis == "All workers",
    Measure == "Hours worked"
  ) %>% 
  select(year = Year, hours = Value) %>% 
  filter(year >= 1948, year <= 2022) %>% 
  mutate(hours = as.numeric(hours)) %>% 
  summarize(hours = mean(hours), .by = year)

productivity <- bea_ndp %>% 
  full_join(bls_hours, by = "year") %>% 
  mutate(prod = ndp / hours)

prod_base <- productivity %>% 
  filter(year == 1967) %>% 
  pull(prod)

mw_data <- productivity %>% 
  add_row(year = 2023) %>% 
  mutate(prod_mw = case_when(
    year == 1968 ~ 1.60,
    year >= 1969 ~ 1.60 * lag(prod) / prod_base 
  )) %>% 
  filter(!is.na(prod_mw))

mw_1968 <- "$1.60"
mw_2023_prod <- mw_data %>% 
  filter(year == 2023) %>% 
  pull(prod_mw) %>% 
  scales::label_dollar()()
```

## effects of $17 in 2028
* `r extract_stat("all", "count_affected")` million workers affected, 
`r extract_stat("all", "share_affected")` of the U.S. workforce
* affected year-round workers earn additional 
`r extract_stat("all", "wage_change_avg_ann")` per year
* generate `r extract_stat("all", "wage_change_total_ann")` billion in higher wages
* share of below poverty workers affected:
`r extract_stat("poverty", "share_affected")`
* share of Black workers affected: 
`r extract_stat("black", "share_affected")`
* share of Latino workers affected: 
`r extract_stat("hispanic", "share_affected")`
* share of affected that are Black/Hispanic female: 
`r extract_stat("black_hispanic_female", "group_share_affected")`
* average annual wage increase for Black workers:
`r extract_stat("black", "wage_change_avg_ann")`
* share of affected that are ages 25-54:
`r extract_stat("prime_age", "group_share_affected")`
* share of affected that are teenagers:
`r extract_stat("teen", "group_share_affected")`
* share of affected that are women:
`r extract_stat("female", "group_share_affected")`
* share of affected that work full-time:
`r extract_stat("full_time", "group_share_affected")`
* share of affected that have some college experience:
`r extract_stat("some_college", "group_share_affected")`
* share of affected that have children:
`r extract_stat("parent", "group_share_affected")`

## other stuff
* Value of 2023 fed min wage had it been indexed to productivity since 1968 = 
`r mw_2023_prod`

